---
title: "linear_regression_lab"
author: "Jose Hernandez"
date: "10/4/2019"
output:
  html_document:
    df_print: paged
---

# Linear Regression

We will be using `tidyverse` tools and `lm` for linear regression examples.

```{r load packages}
library(tidyverse)
```


*Create Fake Data*
```{r built in data}
data(longley)
glimpse(longley)
```

### Longley Dataset

Economics data to use for the regression example.

Plot variables `Emplyed` vs `GNP`.  

`GNP` = 'Gross National Product'
`Employed` = 'Number of people employed'

```{r}
ggplot(longley, aes(x=GNP, y=Employed)) + geom_point()
```

### Run a linear regression on y = `Employed` and x = `GNP`

```{r pressure, echo=FALSE}
fit_model <- lm(Employed~ GNP, longley)
summary(fit_model)
```

```{r}
ggplot(longley, aes(x=GNP, y=Employed)) + geom_point() +
geom_smooth(method = lm, se=F)

```

### Making Prediction

What would you predict the number of employed individuals would be for `GNP = 300` using the formula specification?

*Let's use the model to make many predictions*

```{r}
X <- data.frame(GNP = c(300, 200, 500, 100, 600, 1000))
predict(fit_model, newdata = X)
```

*How can it predict a value not in the data?*



